Serious Titles are Boring: predicting the future of dark matter halos is fun.

### 2019-05-15 PaperText.tex
formatting changes, mostly

% Use vector fonts, so it zooms properly in on-screen viewing software
% Don't change these lines unless you know what you are doing
\usepackage[T1]{fontenc}
\usepackage{ae,aecompl}
\usepackage{color}
\newcommand\khb[1]{{\color{blue}#1}}
\newcommand\edits[1]{{\color{red}#1}}
\newcommand\aab[1]{{\color{green}#1}}

@@

% Don't change these lines
\begin{document}
\label{firstpage}
\pagerange{\pageref{firstpage}--\pageref{lastpage}}
\maketitle

% Abstract of the paper
\begin{abstract}
\khb{The evolution of a dark matter halo in a dark matter only simulation is governed purely by Newtonian gravity, making a clean testbed to determine what halo properties drive its fate. Using machine learning, we predict the survival, mass loss, final position, and merging time of subhalos within a cosmological N-body simulation,
focusing on what instantaneous features of the halo, interaction, and environment matter most.}

%using the ability of these models to accurately make predictions as a proxy for the stochasticity of the underlying subhalo evolution.

Survival is well predicted, with 96.5\% accuracy, given only 3 parameters from the initial interaction. However, the mass loss, final location, and merging times are much more stochastic processes, requiring the definition of accuracy to allow for larger errors in prediction to reach similar accuracy percentages \khb{huh? rephrase after the comma} The scale factor, orbital eccentricity, relative velocity, and the masses of the host and subhalo are the only relevant parameters for determining subhalo evolution. In general, subhalos that enter their hosts at a mid-range of initial scale factor (typically a = .6-.8) are most challenging to make predictions for, across all of our final outcomes. More circular subhalo orbits are also easier to predict, except for in the case of predicting disruption, where the opposite appears to be true. We conclude that the detailed evolution of individual subhalos within N-body simulations is quite stochastic, with implications... \khb{I'd prefer redshift to scale factor, because I think more people use z than a}
\end{abstract}

@@

\khb{looking at the Annual Reviews of Astronomy and Astrophysics can help here -- adding a few to get you started!, but this is definitely not complete}

According to the standard \LambdaCDM model of cosmology, dark matter structures in the universe form hierarchically through series of mergers, with larger halos continuously growing through the accretion of smaller subhalos. Once independent halos themselves, these subhalos sink to the center of their "host" halos, losing mass along their orbits due to tidal effects and dynamical friction. However, as has been shown by modern cosmological N-body simulations (CITE, CITE, CITE), a significant number of such subhalos retain some of their mass, remaining as substructures within their hosts today. The study of these substructures has been fundamental to our understanding of many areas of astrophysics, from large-scale structure (CITE, CITE) to the formation and evolution of galaxies (CITE, CITE), 
\khb{https://www.annualreviews.org/doi/abs/10.1146/annurev-astro-081817-051756, https://www.annualreviews.org/doi/abs/10.1146/annurev-astro-081913-040019}
which relies on both accurate subhalo populations and evolution within these cosmological simulations (CITE, CITE).

\khb{https://www.annualreviews.org/doi/abs/10.1146/annurev-astro-091916-055313?intcmp=trendmd}

\khb{looking at the Annual Reviews of Astronomy and Astrophysics can help here -- adding a few to get you started}

Galaxy formation and evolution is commonly studied through modeling the evolution of the subhalos that these galaxies inhabit, with tools such as semi-analytic models (CITE, CITE). 
\khb{https://www.annualreviews.org/doi/abs/10.1146/annurev-astro-082812-140951?intcmp=trendmd}
These techniques typically rely on merger trees constructed from N-body simulations, along with analytic treatments of the physical processes that cause their evolution (CITE, CITE, CITE). These semi-analytic models often perform quite well, reproducing the statistical properties of subhalo populations (CITE, CITE, CITE). Other works have focused on reproducing specific properties of subhalos, such as disruption rates (CITE, CITE, CITE), mass loss histories (CITE, CITE, CITE), spatial distributions within host halos (CITE, CITE, CITE), and merging timescales (CITE, CITE, CITE).

Although much work has been done to better model these processes and create prescriptions for subhalo evolution, the validity (\textit{doesn't seem like the right word}) of tuning models to the evolution within N-body simulations has remained relatively unexplored. N-body simulations produce consistent subhalo mass functions that have been shown to be converged down to small masses (CITE, CITE, CITE). However, it has not been explored in detail whether the evolution of an individual subhalo within these simulations is a truly deterministic process. These analytic models for subhalo evolution often rely on some set of parameters about the subhalo and host halo at the beginning of an interaction (CITE, CITE, CITE, CITE) to predict the final evolution of such subhalo. However, it may be that subhalo evolution within N-body simulations is a stochastic process, and interactions with similar properties do not necessarily behave in similar ways.

In this work, we use a merger tree generated from the dark matter only simulation VISHNU (CITE?) in order to determine the evolution and fate of subhalos, quantified by the aforementioned final quantities (i.e survival/disruption, mass loss, final location within host halo, merging timescale) using machine learning. We model these final conditions from initial conditions at the time of a subhalo entering its host. Using physically-motivated parameters from the time of the subhalos entry, we use machine learning to predict these final quantities of the subhalo, in the hopes of investigating to what degree subhalo fate is motivated by these parameters and to what degree the interaction is stochastic and cannot be predicted. If the mass loss from a subhalo is deterministic, a machine learning algorithm should be able to map subhalo fate to initial conditions. On the other hand, if there is a level of stochasticity in subhalo fate, whether from physics or from the simulation, there will remain large prediction errors, even when using a complete set of parameters. \textit{This feels very weirdly worded and/or not really concise. Need to revisit.}

Machine learning has emerged as a powerful tool in astrophysics in a variety of fields, such as gravitational wave data analysis, exoplanet detection, galaxy morphological classification, one more. (CITE, CITE, CITE). The ability of machine learning models to approximate any function using a large set of parameters provides a useful means of revealing complicated correlations when a direct analytic function cannot be found. Notably, \citet{Nadler2017} recently used machine learning to predict the survival or disruption of subhalos in a hydrodynamic simulation, using the initial conditions of their counterparts in a dark matter only simulation. They were quite successful, accurately predicting the results of 85\% of their subhalo population. Works like this are encouraging that machine learning can be used to fit these complicated interactions. \edits{However, in this work, although we only use a dark matter only simulation, we also hope to take our predictions a step further by also predicting the mass loss, final spatial position, and merging time of these subhalos.}

\texit{Not positive about this being here or in the section where I talk about the parameters we have? And how much to go through individual papers or just list off like: these are the most common ones for merge time: list, list, list, list (CITE, CITE, CITE, CITE) type deal. So for now, this is just an unfinished paragraph.}
As mentioned above, many works have already attempted to analytically model these subhalo evolution processes with much success. As such, we have selected similar or the same initial parameters that have been found to be important for subhalo evolution to use in our model. For subhalo survival, ....

In Section \ref{sec:simulation}, we describe the simulation and data that was used, as well as the methods to properly reduce the data into the desired sample. In Section \ref{sec:ML Methods}, we cover the machine learning methods used to create our predictive models. In Section \ref{sec:feature selection}, we discuss the feature selection methods used to reduce the number of parameters needed to make predictions. In Section \ref{sec:Results}, we share the results of our models for predicting each of our outcomes, including their performance and which parameters were needed as inputs to the model. Finally, in Section \ref{sec:Conclusion} we discuss implications of the results for both observation and theory.

\section{Description of the Data}
\label{sec:simulation}
\begin{figure}
	% To include a figure from a file named example.*
	% Allowable file formats are eps or ps if compiling using latex
	% or pdf, png, jpg if compiling using pdflatex
	\includegraphics[width=\columnwidth]{Figures/explanatory_figures}
	\vspace{-15pt}
    \caption{An example of a surviving (top) and merging (bottom) interaction between a subhalo and host halo. The orbit of these two subhalos are shown within the radius of their respective host halos, shown by the larger, unfilled circles. Point size corresponds to subhalo mass along the orbit, plotted at each eighth timestep. The green point and circle show initial quantities, at the timestep right before the subhalo enters its host. The purple point and circle show final quantities, at either the timestep right before the subhalo dissolves in the merging case, or at the final timestep in the simulation in the surviving case. Predicted quantities (light blue) are labeled and numbered in the order they will be presented throughout the paper.}
    \label{fig:explanatory_figures}
\end{figure}

	Our analysis makes use of VISHNU, a cosmological N-body simulation with 1000 snapshots for exquisite time resolution; no snapshot is separated by more than X years. VISHNU contains 1680\textsuperscript{3} particles in a volume of 130h\textsuperscript{-1} cMpc and uses WMAP-1 cosmology (\citet{Spergel2003}; $\Omega$\textsubscript{m} = 0.25, $\Omega$\textsubscript{$\Lambda$} = 0.75, $\Omega$\textsubscript{b} = 0.04, $\sigma$\textsubscript{8} = 0.8, \textit{n\textsubscript{s}} = 1.0, \textit{h} = 0.7). Each dark matter particle has mass \textit{m\textsubscript{p}} = 3.215 $\times$ 10\textsuperscript{7}h\textsuperscript{-1} M\textsubscript{\(\odot\)}, and the force resolution is X parsecs. The ROCKSTAR halo finder was used to identify halos and subhalos, and merger trees were constructed with Consistent-Trees (\citet{Behroozi2013}).
    Starting with halos at \textit{z}=0, we use the merger trees to identify the most massive progenitors of all host halos within the simulation, and track the subhalos within. These subhalos are allowed to have further substructure inside of them but cannot, at any point during their infall, become sub-substructure themselves. To help mitigate resolution uncertainties, we select only subhalos with a minimum of 1000 particles (total mass 3.215 $\times$ 10\textsuperscript{10}h\textsuperscript{-1} M\textsubscript{\(\odot\)}) at their time of accretion. We define the accretion time as the last timestep before a subhalo enters its host. 
    %As such, at the initial time of accretion, the will-be subhalos are still host halos themselves. Once the subhalo has entered its host, however, it cannot leave again (so, a flyby interaction would not be considered on its first pass but will be considered on a later pass if it falls back into and remains inside the host). 
    Once it has been accreted, the subhalo must have one of two fates: merge with the host, or remain a bound, identified subhalo within that host until today. We define the point of merging as the point right before dissolution of the subhalo, the last timestep in the simulation where it is identified as its own entity. Following uncertainties in subhalo mass loss shown by \citep{VandenBosch2018}, once a subhalo has lost more than 90\% of its mass, we also consider it to be merged. This cut changes the fates of only a small number of our subhalos, \edits{around 5\%}, but ensures a more consistent definition of merging that is less sensitive to resolution errors. \edits{Counts before/after 1000 particle cut, and before/after this 90\% loss cut is made}.

@@

    In addition to making cuts for resolution, some interactions were removed due to their unphysical behavior, likely as a result of errors in the merger tree generation or halo finding. \textbf{INSERT NUMBER} Subhalos were removed that gained more than 10 times their mass during infall. These subhalos were likely misidentified as they moved close to the center of the host, instead being assigned the mass of another subhalo or the host itself. \textbf{INSERT NUMBER} subhalos were removed because their initial mass was larger than that of its supposed host. \textbf{INSERT NUMBER} subhalos were removed because their positions were outside of the host virial radius. \edits{Actually, are these last two okay because I'm taking the timestep right before entry instead of right after?} \textbf{INSERT NUMBER} subhalos were removed because their time of merging was earlier than their time of entry?? \edits{Need to double check that those cases weren't errors in the writing out of the file or something, I think there were only like 5 so it's weird.} Altogether, we culled about 2\% of the total sample.

@@

    Our resulting sample includes a total of \textbf{INSERT NUMBER} subhalo-halo interactions, with \textbf{INSERT NUMBER} of such being mergers and \textbf{INSERT NUMBER} being survivors. Distributions of this sample with respect to host masses, mass ratios, and the scale factor of the time of entry are shown in Fig.~\ref{fig:combined_distributions_logBin}. The most common interactions in our sample are mergers of unequal masses that occurred more recently. However, our sample also spans the space of more equal mass and higher redshift mergers, with hundreds of interactions shown in many of the bins in Fig.~\ref{fig:combined_distributions_logBin}. The effects of our chosen particle cut can also be clearly seen in Fig.~\ref{fig:combined_distributions_logBin}. Because subhalos must have 1000 particles at their time of entry, this results in a minimum initial mass of 3.215 $\times$ 10\textsuperscript{10}h\textsuperscript{-1} M\textsubscript{\(\odot\)} for both the subhalo and host halo, given that a host halo must also be at least as large as its subhalo. In the left panel of Fig.~\ref{fig:combined_distributions_logBin}, this results in an area of no data with low mass hosts and unequal masses, because most subhalos of low mass hosts are too low mass to be included in our sample. \edits{Say something about if these counts are typical/ what we expect to see/ agreement with other simulations?}
\par
    For both the host halo and subhalo at the timestep of infall, and at either the timestep right before merging or at z=0, we take several physically-motivated parameters to describe the interaction between the halos. A few additional parameters were not given explicitly from Consistent-Trees merger trees, but were calculated from some set of given parameters. We divide our final set of initial parameters, which were used as input parameters to the machine learning models to make predictions, into four main categories. Internal halo parameters are those that are properties of the individual halos themselves. These parameters do not depend on the other halo in the interaction, instead giving important information about the structure of the halos as individuals. Relative halo parameters describe properties of the halos that are dependent on either other halos within the simulation or the simulation itself. Orbital parameters provide information about the initial trajectory of the subhalo's infall path. Environmental parameters describe the influence of larger scale environment around the subhalo and its host. In total, this gives us 26 input parameters, with definitions:

@@

\khb{merger and accretion can be conflated, so maybe use dissolves or something to clarify 'merger'}

@@

        \item [1.] \textbf{survival}: a 0 or 1, depending on if the subhalo exists above the required mass threshold at z=0 (survives, 1) or if the subhalo has merged with its host at some time before (merges, 0). Predicted for all subhalos.
        \item [2.] \textbf{M\textsubscript{sub,f}}: the virial mass of the subhalo at the end of the interaction. Only predicted for subhalos that survive until z=0.

@@

        \item [4.] \textbf{t\textsubscript{merge}}: the elapsed time between entry of a subhalo into the host virial radius and the time of merging with the host. Only predicted for subhalos that merge at some time before z=0.

@@

    As preparation for using machine learning, we split the data into \textit{train}  and \textit{test} sets, with 80\% of the data for training and 20\% to test. These subsamples are selected randomly from the total sample. The data are scaled and normalized to be distributed as a Gaussian with zero mean and unit variance, meaning that before giving each parameter to the model, we transform it by: 

@@

    Each time we train our machine learning models, we further divide the training set into training and validation sets, also using an 80/20 split. From the original dataset, this means that 20\% of our data is used only for final testing, while 80\% of our data set is repeatedly divided into training and validation sets. The use of these validation sets allows us to tune model hyperparameters while ensuring that the testing set is not overfit.

\section{Machine Learning Methods}
\label{sec:ML Methods}
The machine learning algorithms used in this work come from the \texttt{scikit-learn} package for python. Subhalo survival can be thought of as a classification problem, which is well-suited to a random forest algorithm. On the other hand, predictions of the amount of mass loss, the final position, and merging time are classic regression problems, and we use the gradient boosting regressor algorithm. Although we refer the reader to the \texttt{scikit-learn} documentation for a full description of these algorithms, we briefly describe these methods below.

\khb{general point: please switch to active voice}

>>>>

% Use vector fonts, so it zooms properly in on-screen viewing software
% Don't change these lines unless you know what you are doing
\usepackage[T1]{fontenc}
\usepackage{ae,aecompl}
\usepackage{color}
\newcommand\khb[1]{{\color{magenta}#1}}
\newcommand\edits[1]{{\color{red}#1}}
\newcommand\aab[1]{{\color{green}#1}}

@@

% Don't change these lines

\hypersetup{draft} %% ABBIE DELETE THIS LATER THIS IS BAD
\begin{document}
\label{firstpage}
\pagerange{\pageref{firstpage}--\pageref{lastpage}}
\maketitle

% Abstract of the paper
\begin{abstract}
The evolution of a dark matter halo in a dark matter only simulation is governed purely by Newtonian gravity, making a clean testbed to determine what halo properties drive its fate. Using machine learning, we predict the survival, mass loss, final position, and merging time of subhalos within a cosmological N-body simulation, focusing on what instantaneous initial features of the halo, interaction, and environment matter most. Survival is well predicted, with our model achieving 96.5\% accuracy using only 3 parameters from the initial interaction. However, the mass loss, final location, and merging times are much more stochastic processes, with significant margins of error between the true and predicted quantities for much of our sample. The redshift, orbital eccentricity, relative velocity, and the masses of the host and subhalo are the only relevant initial parameters for determining subhalo evolution. In general, subhalos that enter their hosts at a mid-range of redshifts (typically z = .67-.43) are the most challenging to make predictions for, across all of our final outcomes. More circular subhalo orbits are also easier to predict, except for in the case of predicting disruption, where the opposite appears to be true. We conclude that the detailed evolution of individual subhalos within N-body simulations is quite difficult to predict, pointing to a stochasticity in the merging process. We discuss implications for both simulations and observations.
\end{abstract}

@@

According to the standard \textLambda CDM model of cosmology, dark matter structures in the universe form hierarchically through series of mergers, with larger halos continuously growing through the accretion of smaller subhalos. Once independent halos themselves, these subhalos sink to the center of their "host" halos, losing mass to their hosts along their orbits due to tidal effects and dynamical friction, the effects of which have been studied in detail (\citet{Tormen1998}, \citet{Weinberg1989}, \citet{Bosch1999}, \citet{Hayashi2003}, \citet{Taffoni2003}, \citet{Gan2010}, \citet{VandenBosch2017}). However, as has been shown by modern cosmological N-body simulations, a significant number of such subhalos retain some of their mass, surviving as substructures within their hosts today. The study of these substructures has been fundamental to our understanding of many areas of astrophysics, from large-scale structure (\citet{Knebe2004}, \citet{Zentner2003}) to the formation and evolution of galaxies (\citet{Hayashi2009}, \citet{Kazantzidis2009}, \citet{Simha2016}), which rely on both accurate final subhalo populations and the evolution of these populations within these cosmological simulations (\citet{Diemand2007}, \citet{Giocoli2007}).

Galaxy formation and evolution is commonly studied through modeling the evolution of the subhalos that these galaxies inhabit, with tools such as semi-analytic models (\citet{Taylor2003}, \citet{Zentner2004}, \citet{Penarrubia2005}, \citet{Jiang2016}). These techniques typically rely on merger trees constructed from N-body simulations, along with analytic treatments of the physical processes that cause their evolution. These semi-analytic models often perform quite well, reproducing the statistical properties of subhalo populations. Other works have focused on reproducing specific properties of subhalos, such as disruption rates, mass loss histories, spatial distributions within host halos, and merging timescales.

Although much work has been done to better model these processes and create prescriptions for subhalo evolution, the reliability of tuning models to the evolution within N-body simulations has remained relatively unexplored. N-body simulations produce consistent subhalo mass functions and the evolution of these populations has been thoroughly studied (\citet{Gao2004}, \citet{Onions2012}, \citet{Jiang2016a}, \citet{Chua2016}) and much recent work has been done to better constrain the low mass end of the subhalo mass function to increase our understanding of smaller satellites (\citet{Munshi2018}). However, it has not been explored in detail whether the evolution of an individual subhalo within these simulations is a truly deterministic process. Models for subhalo evolution that are tuned to the evolution within N-body simulations (\citet{Penarrubia2005}, \citet{Gan2010}, \citet{Hiroshima2018}) reflect the noise and uncertainty in the merging process within the simulation. It may be that subhalo evolution within N-body simulations is somewhat stochastic, such that interactions with similar properties do not necessarily behave in similar ways, adding unforseen complications when using these models.

In this work, we use a merger tree generated from the dark matter only simulation VISHNU, available on the the Theoretical Astrophysical Observatory (\citet{Bernyk2014}), in order to determine the evolution and fate of subhalos, quantified by the aforementioned final quantities (i.e survival/disruption, mass loss, final location within host halo, merging timescale) using machine learning. We model these final conditions from initial conditions at the time of a subhalo entering its host. Using physically-motivated parameters from the time of the subhalos entry, we use machine learning to predict these final quantities of the subhalo, in the hopes of investigating to what degree subhalo fate is motivated by these parameters and to what degree the interaction is stochastic and cannot be predicted. If the mass loss from a subhalo is deterministic, a machine learning algorithm should be able to successfully map subhalo initial conditions to their fates. On the other hand, if there is a level of stochasticity in subhalo fate, there will remain large prediction errors, even when using a complete set of parameters.

Machine learning has emerged as a powerful tool in astrophysics with a variety of applications, such as galaxy classification (\citet{Barchi2019MachineCatalog}, \citet{Nolte2019GalaxyData}), exoplanet detection (\citet{Schanche2018Machine-learningSurveys}), and gravitational wave noise removal (\citet{Cavagli2018FindingLearning}), and has recently been used with cosmological simulations to predict galaxy properties from halo properties (\citet{Kamdar2016}), populate halos with galaxies (\citet{Agarwal2017}), connect initial conditions to final halos (\citet{Lucie-Smith2018}), and predict the halo masses of galaxies (\citet{Calderon2019PredictionApproach}). The ability of machine learning models to approximate any function using a large set of parameters provides a useful means of revealing complicated correlations when a direct analytic function cannot be found. Notably, \citet{Nadler2017} recently used machine learning to predict the survival or disruption of subhalos in a hydrodynamic simulation, using the initial conditions of their counterparts in a dark matter only simulation. They were quite successful, accurately predicting the results of 85\% of their test set of subhalos. Works like this are encouraging that machine learning can be used to fit these complicated interactions.

In this work, we instead focus on dark matter only simulations, where we predict not only survival, but more quantitative metrics such as the amount of mass loss, the final position, and the duration of the merging time for a subhalo. By using a complete set of physical parameters to make these predictions, we hope to reveal what subhalo parameters are the most closely tied to -- and thus what physical processes most strongly drive -- this evolution. While we expect these additional quantities to be more difficult to predict, even in a dark-matter only simulation with simpler physics, than a binary prediction for disruption, the ability or inability of machine learning models to make predictions in the first place can inform us about the determinism of a model. Predictions that are not successful, despite having this complete parameter set available to them, may point to problems in the assumption that subhalos evolve in a predictable way.

As mentioned, many works have already attempted to either model these subhalo evolution processes from well-understood physics, or investigate trends in subhalo populations within simulations to better understand what drives subhalo evolution. As such, we look to these works to select similar or the same initial parameters that have been found to be important for subhalo evolution to use in our model. The topic of subhalo disruption has a particularly rich body of work. Several works (\citet{Ghigna2000}, \citet{Diemand2004}, \citet{Gao2004}, \citet{Zentner2004}, \citet{Penarrubia2005}, \citet{Diemand2007}) have found that redshift of entry is overwhelmingly important in determining the survivability of subhalos, finding that the majority of surviving subhalos were accreted more recently than z=1. \citet{Gao2004}, \citet{Giocoli2009}, and \citet{Gao2011} find that the abundance of subhalos is lower for host halos of fixed mass that have higher concentrations, but note that higher concentration halos on average form earlier and thus would have less surviving substructure because their substructure is accreted earlier and spends more time orbiting inside the host. \citet{Klimentowski2010} find that many subhalos that are destroyed do not complete even one pericenter passage, but many subhalos that do survive have completed more than one pericenter passage. They also find that surviving subhalos tend to have eccentric orbits. \citet{Diemand2004} find that slower subhalos with low orbital energies were preferentially destroyed, resulting in a positive velocity bias in the subhalo distribution within clusters. \citet{Yang2009} model the fraction of surviving subhalos as a function of the subhalo to host halo mass ratio, with good results. \citet{Tormen1998} find similar results, with larger mass ratios leading to more rapid disruption. \citet{Gill2004} find only weak dependence of the subhalo disruption rate on the mass of the host halo.

As subhalos typically disrupt after losing a significant amount of their mass, we expect many of the parameters that determine subhalo survival to also be of significant importance to subhalo mass loss. Additionally, some works have focused specifically on the topic of subhalo mass loss. \citet{Gao2004} look at populations of subhalos in a wide variety of host halos and find that subhalo mass loss does not strongly depend on the mass of the host halo. \citet{Taylor2003} model subhalo mass loss in a static host potential and find that the eccentricity of the subhalo orbit and the subhalo concentration are the dominant determinants of mass loss, with subhalos losing a significant portion of their mass during each pericenter passage, resulting in more radial orbits losing mass more quickly. \citet{VanDenBosch2005} look at the average mass loss rates of subhalos using only the redshift and subhalo to host halo mass ratio, and find good agreement with subhalo mass functions from simulations. Several additional works (\citet{Taylor2001}, \citet{Hayashi2003}, \citet{Kampakoglou2007}, \citet{Gan2010}, \citet{Han2016}) have modeled subhalo positions and internal structure at each timestep to analytically model their evolution. Although these works give deeper insights to the relative importance of the physical processes at work on these subhalos, our machine learning models do not explicitly model the evolution over time of our subhalos, so the parameters used by those works are less relevant here.

The final distributions of subhalos within their hosts have also been closely studied. \citet{Angulo2009} find that the radial distributions of subhalos does not depend on host halo mass or redshift, but does depend on the suhalo to host halo mass ratio. Similarly, \citet{Nipoti2018} find that subhalos that merge with the central parts of their host halos tend to have larger subhalo to host halo mass ratios than those merging with the outer parts of the host halo, which may suggest that mass ratio can help predict the final location of a subhalo. \citet{Reed2005} find a trend of decreasing subhalo spin closer to the host center, suggesting that lower spin halos may have more success surviving to z=0 when orbiting at small fractions of the host radius. If this is due to higher spin subhalos being more susceptible to tidal stripping, this trend could also be important for our other predicted quantities.

Analytic predictions of merging timescales for subhalos have been found by a number of previous works. \citet{Wetzel2010} fit a function to determine the time until satellite removal after accretion using only the subhalo to host halo mass ratio. \citet{Boylan-Kolchin2007} and \citet{Jiang2007} perform similar studies, modeling the effects of dynamical friction to determine galaxy merging timescales, using the subhalo to host halo mass ratio, the circularity and energy of the subhalo orbit, and the virial radius of the host halo, and the dynamical time at the host halo's virial radius. \citet{McCavana2012} model merging timescales using similar parameters. Although they find different dependencies on these parameters than \citet{Boylan-Kolchin2007} or \citet{Jiang2007}, the parameters dominating the merging timescale remain the same.

In Section \ref{sec:simulation}, we describe the simulation and data that we use, as well as the methods we employ to properly reduce the data into our desired sample. In Section \ref{sec:ML Methods}, we cover the machine learning methods we use to create our predictive models. In Section \ref{sec:feature selection}, we discuss the feature selection methods we use to gain intuition and decide on which parameters are the most important for each model to make predictions. In Section \ref{sec:Results}, we share the results of our models for predicting each of our outcomes, including their performance and which parameters were needed as inputs to the model. Finally, in Section \ref{sec:Conclusion} we discuss implications of our results for both observation and theory.

\section{Description of the Data}
\label{sec:simulation}
\begin{figure}
	% To include a figure from a file named example.*
	% Allowable file formats are eps or ps if compiling using latex
	% or pdf, png, jpg if compiling using pdflatex
	\includegraphics[width=\columnwidth]{Figures/explanatory_figures}
	\vspace{-15pt}
    \caption{An example of a surviving (top) and merging (bottom) interaction between a subhalo and host halo. The orbit of these two subhalos are shown within the radius of their respective host halos, shown by the larger, unfilled circles. Point size corresponds to subhalo mass along the orbit, plotted at each eighth timestep. The green point and circle show initial quantities, at the timestep right before the subhalo enters its host. The orange point and circle show final quantities, at either the timestep right before the subhalo dissolves in the merging case, or at the final timestep in the simulation in the surviving case. Predicted quantities (gold) are labeled and numbered in the order we will present them throughout the paper.}
    \label{fig:explanatory_figures}
\end{figure}

Our analysis makes use of VISHNU, a cosmological N-body simulation with 1000 snapshots for exquisite time resolution; no snapshot is separated by more than 3.2 $\times$ 10\textsuperscript{7} years. VISHNU contains 1680\textsuperscript{3} particles in a volume of 130h\textsuperscript{-1} cMpc and uses WMAP-1 cosmology (\citet{Spergel2003}); $\Omega$\textsubscript{m} = 0.25, $\Omega$\textsubscript{$\Lambda$} = 0.75, $\Omega$\textsubscript{b} = 0.04, $\sigma$\textsubscript{8} = 0.8, \textit{n\textsubscript{s}} = 1.0, \textit{h} = 0.7). Each dark matter particle has mass \textit{m\textsubscript{p}} = 3.215 $\times$ 10\textsuperscript{7}h\textsuperscript{-1} M\textsubscript{\(\odot\)}, and the force resolution is \edits{\textbf{INSERT NUMBER} (2.2 kpc?)} parsecs. The ROCKSTAR halo finder was used to identify halos and subhalos (\citet{Behroozi2013}), and merger trees were constructed with Consistent-Trees (\citet{Behroozi2013a}).

Starting with halos at \textit{z}=0, we use the merger trees to identify the most massive progenitors of all host halos within the simulation, and track the subhalos within. These subhalos are allowed to have further substructure inside of them but cannot, at any point during their infall, become sub-substructure themselves. To help mitigate resolution uncertainties, we select only subhalos with a minimum of 1000 particles (total mass 3.215 $\times$ 10\textsuperscript{10}h\textsuperscript{-1} M\textsubscript{\(\odot\)}) at their time of accretion. We define the accretion time as the last timestep before a subhalo enters its host. This mass cut reduces our sample from 1284058 subhalos to 121343 subhalos.

Once it has been accreted, the subhalo must have one of two fates: merge with the host, or remain a bound, identified subhalo within that host until today. We define the point of merging as the point right before dissolution of the subhalo, the last timestep in the simulation where it is identified as its own entity. Following uncertainties in subhalo mass loss shown by \citet{Bosch2018}, once a subhalo has lost more than 90\% of its mass, we also consider it to be merged. This cut changes the fates of 21929 subhalos, around 18\% of our total sample, but ensures a more consistent definition of merging that is less sensitive to resolution errors.

@@

In addition to making cuts for resolution, some interactions were removed due to their unphysical behavior, likely as a result of errors in the merger tree generation or halo finding. 1474 subhalos were removed that more than tripled their mass during infall. These subhalos were likely misidentified as they moved close to the center of the host, instead being assigned the mass of another subhalo or the host itself. 327 subhalos were removed because their initial mass was larger than that of their supposed hosts. Altogether, we culled about an additional 1.5\% of the total sample with these cuts.

@@

    Our resulting sample includes a total of 119543 subhalo-halo interactions, with 76442 of such being mergers and 43101 being survivors. Distributions of this sample with respect to host masses, mass ratios, and the scale factor of the time of entry are shown in Figure~\ref{fig:combined_distributions_logBin}. The most common interactions in our sample are mergers of unequal masses that occurred more recently. However, our sample also spans the space of more equal mass and higher redshift mergers, with hundreds of interactions shown in many of the bins in Figure~\ref{fig:combined_distributions_logBin}. The effects of our chosen particle cut can also be clearly seen in Figure~\ref{fig:combined_distributions_logBin}. Because subhalos must have 1000 particles at their time of entry, this results in a minimum initial mass of 3.215 $\times$ 10\textsuperscript{10}h\textsuperscript{-1} M\textsubscript{\(\odot\)} for both the subhalo and host halo, given that a host halo must also be at least as large as its subhalo. In the left panel of Figure~\ref{fig:combined_distributions_logBin}, this results in an area of no data with low mass hosts and unequal masses, because most subhalos of low mass hosts are too low mass to be included in our sample.
\par
    For both the host halo and subhalo at the timestep of infall, and at either the timestep right before merging or at z=0, we take several physically-motivated parameters to describe the interaction between the halos. We also calculate a few additional parameters that were not given explicitly from Consistent-Trees merger trees. We divide our final set of initial parameters, which were used as input parameters to the machine learning models to make predictions, into four main categories. Internal halo parameters are those that are properties of the individual halos themselves. These parameters do not depend on the other halo in the interaction, instead giving important information about the structure of the halos as individuals. Relative halo parameters describe properties of the halos that are dependent on either other halos within the simulation or the simulation itself. Orbital parameters provide information about the initial trajectory of the subhalo's infall path. Environmental parameters describe the influence of larger scale environment around the subhalo and its host. In total, this gives us 26 input parameters, with definitions:

@@

@@

        \item [1.] \textbf{survival}: a 0 or 1, depending on if the subhalo exists above the required mass threshold at z=0 (survives, 1) or if the subhalo has either merged with its host or fallen below the mass threshold at some time before z=0 (merges, 0). Predicted for all subhalos.
        \item [2.] \textbf{M\textsubscript{sub,f}}: the M\textsubscript{200\textsubscript{b}} of the subhalo at the end of the interaction. Only predicted for subhalos that survive until z=0.

@@

        \item [4.] \textbf{t\textsubscript{merge}}: the elapsed time between accretion of a subhalo into the host radius and the time of merging with the host. Only predicted for subhalos that dissolve at some time before z=0.

@@

    As preparation for using machine learning, we split the data into \textit{train}  and \textit{test} sets, with 80\% of the data for training and 20\% to test. These subsamples are selected randomly from the total sample. The data are scaled and normalized to be distributed as a Gaussian with zero mean and unit variance, meaning that before giving each parameter (X) to the model, we transform it by: 

@@

    Each time we train our machine learning models, we further divide the training set into training and validation sets, also using an 80\%/20\% split. From the original dataset, this means that 20\% of our data is used only for final testing, while 80\% of our data set is repeatedly divided into training and validation sets. The use of these validation sets allows us to tune model hyperparameters while leaving the testing set untouched to ensure that it cannot be overfit.

\section{Machine Learning Methods}
\label{sec:ML Methods}
The machine learning algorithms we use in this work come from the \texttt{scikit-learn} package for python. Subhalo survival is a classification problem, so we use a random forest algorithm. On the other hand, predictions of the amount of mass loss, the final position, and merging time are all classic regression problems, so we use the gradient boosting regressor algorithm. Although we refer the reader to the \texttt{scikit-learn} documentation for a full description of these algorithms, we briefly describe these methods, and others that we tried, below.

\subsection{Random Forests}
\label{sec:rf}
Random forest classifiers use a compilation of decision trees to reach consensus on a prediction. These decision trees repeatedly split the data on the values of its input parameters, binning the data into small subsets until bins are small enough that they generally give the correct output. In a random forest, many individual decision trees are trained on random subsets of the full dataset. Because random forests are a type of bagging - or bootstrap aggregating - method, the classification for an object is then the majority vote of all of the trees. This means that the trees that make up the ensemble are individual from each other - each tree makes an independent prediction for the classification of an object in the testing set.

There are several hyperparameters of the algorithm that we tune in order to get the best-fitting model. The \textit{n\_estimators} hyperparameter sets the number of estimators, in this case decision trees, that will be separately trained and used in the final consensus. Using too few decision trees removes the power of using multiple trees. In general, using too many trees is not a concern, as the performance of random forests is not decreased by adding large numbers of trees. For much larger datasets, adding too many trees can lead to much longer training times. However, given our relatively small number of samples, this isn't much of a concern in our training.  The \textit{max\_depth} hyperparameter sets the maximum number of decisions that can be made in each tree before reaching a classification decision. Effectively, this sets the maximum number of input parameters each decision can use, as at each depth level of a decision tree, an input can only follow one path, splitting on one parameter. The \textit{max\_leaf\_nodes} hyperparameter sets the maximum number of new nodes a parameter can be divided into from a single leaf, which is one single path that a decision has taken at a given depth level. We note that, if this value is too small, the decision tree may need to split on the same input parameter at multiple depth levels. We keep other parameters of the algorithm set to their default values, which can be found in the \texttt{scikit-learn} documentation. In general, these other hyperparameters tend to allow the user to fine-tune the criterion for the decision tree to make a split, or add criterion to prevent overfitting, both of which we did not find necessary when creating our model.
